---
title: "Alpha/Beta Diversities"
author: 
  - Huang Lin$^1$
  - $^1$Department of Biostatistics, University of Pittsburgh, 130 De Soto Street, Pittsburgh, PA 15261 USA
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document: 
    toc: true
    theme: united
---

```{r setup, include=FALSE}
# rm(list = ls())

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      message = FALSE, comment = NA, 
                      fig.width = 6.25, fig.height = 5)

library(readr)
library(openxlsx)
library(tidyverse) 
library(microbiome)
library(vegan)
library(compositions)
library(magrittr)
library(qwraps2)
library(knitr)
library(kableExtra)
options(digits = 2)
```

```{r data}
# OTU table
otu_table = read_tsv("../data/otu-table.tsv", skip = 1)
otu_id = otu_table$`#OTU ID`
otu_table = data.frame(otu_table[, -1], check.names = FALSE, row.names = otu_id)

# Taxonomy table
tax = read_tsv("../data/taxonomy.tsv")
otu_id = tax$`Feature ID`
tax = data.frame(tax[, - c(1, 3)], row.names = otu_id)
tax = tax %>% separate(col = Taxon, 
                       into = c("Kingdom", "Phylum", "Class", "Order", 
                                "Family", "Genus", "Species"),
                       sep = ";")
for (i in 1:ncol(tax)) {
  tax[, i] = sapply(tax[, i], function(x) str_split(x, "__")[[1]][2])
}
tax = as.matrix(tax)
tax[tax == ""] = NA

# Tree
tree = read_tree("../data/tree.nwk")

# Meta data
meta_data = read_tsv("../data/metadata.tsv")
meta_data = meta_data %>%
  mutate_if(is.character, as.factor)
meta_data$sampleid = as.character(meta_data$sampleid)
meta_data$caregiver_stress_level = factor(meta_data$caregiver_stress_level,
                                          levels = c("Low",  "Medium",  "High"))
meta_data$depression_level = factor(meta_data$depression_level,
                                    levels = c("Low",  "Medium",  "High"))
meta_data$hostility_level = factor(meta_data$hostility_level,
                                   levels = c("Low",  "Medium",  "High"))
meta_data$das_level = factor(meta_data$das_level,
                             levels = c("Low",  "Medium",  "High"))
meta_data$metabolic_syndrome_level = factor(meta_data$metabolic_syndrome_level,
                                            levels = c("Low",  "Medium",  "High"))

OTU = otu_table(otu_table, taxa_are_rows = TRUE)
META = sample_data(meta_data)
sample_names(META) = meta_data$sampleid
TAX = tax_table(tax)
otu_data = phyloseq(OTU, TAX, META, tree)
```

# 1. Analyses at genus level

```{r}
# Aggregate taxa
genus_data = aggregate_taxa(otu_data, "Genus")
genus_data2 = merge_taxa2(genus_data, pattern = "\\Clostridium", name = "Clostridium")
genus_rarefied = rarefy_even_depth(genus_data2, rngseed = 1, 
                                   sample.size = 0.9 * min(sample_sums(genus_data2)), 
                                   replace = FALSE)
```

## 1.1 Alpha diversity: Shannon’s diversity index

*P-value is obtained by Kruskal-Wallis Rank Sum Test.*

```{r}
d_alpha = alpha(genus_rarefied, index = "diversity_shannon")
df_alpha = data.frame(d = d_alpha$diversity_shannon,
                      meta(genus_rarefied) %>% 
                        dplyr::select(alcohol, 
                                      caregiver_stress_level:metabolic_syndrome_level))
covariates = colnames(df_alpha)[-1]
labels = c("Alcohol Use", "Caregiver Stress Level", "Depression Level",
           "Hostility Level", "DAS Level", "Metabolic Syndrome Level")

for (i in seq_along(covariates)) {
  cat("\n \n \n")
  covariate = covariates[i]
  label = labels[i]
  tformula = formula(paste0("d ~ ", covariate)) 
  p_val = kruskal.test(tformula, data = df_alpha)$p.value
  df_ann = data.frame(x = 2.2, y = 1.01 * max(df_alpha$d), p = p_val)
  df_ann = df_ann %>% 
    mutate(label = paste0("p = ", signif(p, 2)))
  
  df_fig = df_alpha %>% 
    transmute(x = !!as.name(covariate), y = d)
  df_fig = df_fig[complete.cases(df_fig), ]
  
  p = df_fig %>%
    ggplot(aes(x = x, y = y)) + 
    geom_violin(trim = FALSE, aes(fill = x)) +
    geom_boxplot(width = 0.1, fill = "white") +
    scale_fill_discrete(name = NULL) +
    geom_label(data = df_ann, aes(x = x, y = y, label = label),
               size = 4, vjust = -0.5, hjust = 0, color = "orange") +
    labs(x = NULL, y = "Shannon’s diversity index", title = label) + 
    theme_bw() + 
    theme(strip.background = element_rect(fill = "white"), 
          legend.position = "bottom",
          plot.title = element_text(hjust = 0.5))
  print(p)
  cat("\n \n \n")
  fig_path1 = paste0("../images/diversity/genus_alpha_", covariate, ".jpeg")
  fig_path2 = paste0("../images/diversity/genus_alpha_", covariate, ".pdf")
  ggsave(filename = fig_path1, plot = p, height = 5, width = 6.25, 
         units = 'in', dpi = 300) 
  ggsave(filename = fig_path2, plot = p, height = 5, width = 6.25) 
}
```

## 1.2 Beta diversity: Bray-Curtis dissimilarity

*P-values are obtained by Permutational Multivariate Analysis of Variance (PERMANOVA) and Permutational Analysis of Multivariate Dispersion (PERMDISP).*

```{r}
for (i in seq_along(covariates)) {
  cat("\n \n \n")
  covariate = covariates[i]
  label = labels[i]
  pseq = genus_rarefied
  sample_data(pseq)$covariate = meta(pseq)[, covariate]
  pseq_subset = subset_samples(pseq, !is.na(covariate))
  
  set.seed(123)
  # PERMANOVA
  permanova = adonis(t(abundances(pseq_subset)) ~ covariate, 
                     data = meta(pseq_subset), 
                     permutations = 999, method = "bray")$aov.tab
  # PERMDISP
  dis = vegdist(t(abundances(pseq_subset)), method = "bray")
  groups = sample_data(pseq_subset)$covariate
  mod = betadisper(d = dis, group = groups, type = "median")

  # Draw the Plot
  labs = paste0("PCoA", 1:2, " (", signif(100 * mod$eig / sum(mod$eig), 3), "%)")
  # brewer.pal(n = 8, name = "Accent")
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  cat("\n \n \n")
  # Export
  fig_path1 = paste0("../images/diversity/genus_beta_", covariate, ".jpeg")
  jpeg(filename = fig_path1, height = 5, width = 6.25, res = 300, units = "in")
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  dev.off()
  
  fig_path2 = paste0("../images/diversity/genus_beta_", covariate, ".pdf")
  pdf(file = fig_path2, height = 5, width = 6.25)
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  dev.off()
}
```

# 2. Analyses at species level

```{r}
# Aggregate taxa
species_data = aggregate_taxa(otu_data, "Species")
species_rarefied = rarefy_even_depth(species_data, rngseed = 1, 
                                     sample.size = 0.9 * min(sample_sums(species_data)), 
                                     replace = FALSE)
```

## 2.1 Alpha diversity: Shannon’s diversity index

*P-value is obtained by Kruskal-Wallis Rank Sum Test.*

```{r}
d_alpha = alpha(species_rarefied, index = "diversity_shannon")
df_alpha = data.frame(d = d_alpha$diversity_shannon,
                      meta(species_rarefied) %>% 
                        dplyr::select(alcohol, 
                                      caregiver_stress_level:metabolic_syndrome_level))
covariates = colnames(df_alpha)[-1]
labels = c("Alcohol Use", "Caregiver Stress Level", "Depression Level",
           "Hostility Level", "DAS Level", "Metabolic Syndrome Level")

for (i in seq_along(covariates)) {
  cat("\n \n \n")
  covariate = covariates[i]
  label = labels[i]
  tformula = formula(paste0("d ~ ", covariate)) 
  p_val = kruskal.test(tformula, data = df_alpha)$p.value
  df_ann = data.frame(x = 2.2, y = 1.01 * max(df_alpha$d), p = p_val)
  df_ann = df_ann %>% 
    mutate(label = paste0("p = ", signif(p, 2)))
  
  df_fig = df_alpha %>% 
    transmute(x = !!as.name(covariate), y = d)
  df_fig = df_fig[complete.cases(df_fig), ]
  
  p = df_fig %>%
    ggplot(aes(x = x, y = y)) + 
    geom_violin(trim = FALSE, aes(fill = x)) +
    geom_boxplot(width = 0.1, fill = "white") +
    scale_fill_discrete(name = NULL) +
    geom_label(data = df_ann, aes(x = x, y = y, label = label),
               size = 4, vjust = -0.5, hjust = 0, color = "orange") +
    labs(x = NULL, y = "Shannon’s diversity index", title = label) + 
    theme_bw() + 
    theme(strip.background = element_rect(fill = "white"), 
          legend.position = "bottom",
          plot.title = element_text(hjust = 0.5))
  print(p)
  cat("\n \n \n")
  fig_path1 = paste0("../images/diversity/species_alpha_", covariate, ".jpeg")
  fig_path2 = paste0("../images/diversity/species_alpha_", covariate, ".pdf")
  ggsave(filename = fig_path1, plot = p, height = 5, width = 6.25, 
         units = 'in', dpi = 300) 
  ggsave(filename = fig_path2, plot = p, height = 5, width = 6.25) 
}
```

## 2.2 Beta diversity: Bray-Curtis dissimilarity

*P-values are obtained by Permutational Multivariate Analysis of Variance (PERMANOVA) and Permutational Analysis of Multivariate Dispersion (PERMDISP).*

```{r}
for (i in seq_along(covariates)) {
  cat("\n \n \n")
  covariate = covariates[i]
  label = labels[i]
  pseq = species_rarefied
  sample_data(pseq)$covariate = meta(pseq)[, covariate]
  pseq_subset = subset_samples(pseq, !is.na(covariate))
  
  set.seed(123)
  # PERMANOVA
  permanova = adonis(t(abundances(pseq_subset)) ~ covariate, 
                     data = meta(pseq_subset), 
                     permutations = 999, method = "bray")$aov.tab
  # PERMDISP
  dis = vegdist(t(abundances(pseq_subset)), method = "bray")
  groups = sample_data(pseq_subset)$covariate
  mod = betadisper(d = dis, group = groups, type = "median")

  # Draw the Plot
  labs = paste0("PCoA", 1:2, " (", signif(100 * mod$eig / sum(mod$eig), 3), "%)")
  # brewer.pal(n = 8, name = "Accent")
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  cat("\n \n \n")
  # Export
  fig_path1 = paste0("../images/diversity/species_beta_", covariate, ".jpeg")
  jpeg(filename = fig_path1, height = 5, width = 6.25, res = 300, units = "in")
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  dev.off()
  
  fig_path2 = paste0("../images/diversity/species_beta_", covariate, ".pdf")
  pdf(file = fig_path2, height = 5, width = 6.25)
  plot(mod, pch = 15:16, cex.lab = 1.25, cex = 1, 
       main = label, 
       xlab = labs[1], ylab = labs[2], ylim = c(-0.4, 0.6), xaxt = "n",
       col = c("#7FC97F", "#BEAED4"), sub = NULL,
       hull = FALSE, ellipse = TRUE, conf = 0.68) # 68% data coverage for data ellipses
  axis(1, at = round(seq(-0.6, 0.6, by = 0.2), 1), las = 1)
  legend(x = 0.5, y = 0.2, legend = unique(groups),
         col = c("#7FC97F", "#BEAED4"), pch = 15:16, cex = 0.8)
  legend(x = 0.2, y = 0.6, cex = 0.7,
         legend = c(paste0("p (PERMANOVA) = ", signif(permanova$`Pr(>F)`[1], 2)),
                    paste0("p (PERMDISP) = ", signif(permutest(mod)$tab$`Pr(>F)`[1], 2))))
  dev.off()
}
```

# Session information

```{r, message = FALSE, warning = FALSE, comment = NA}
sessionInfo()
```







